---
title: "R4DS Thorndike pages 28 - 38"
author: "Hoffman"
format: 
  revealjs:
    echo: true
---

## Thorndike 28-38

In this presentation, we will cover pages 28-38 of Thorndike and Thorndike-Christ (2009). 

First, load the necessary libraries and data.

```{r}
library(tidyverse)
library(googlesheets4)
gs4_deauth() # deauthorize Google Sheet so that anyone can access it
students <- read_sheet("https://docs.google.com/spreadsheets/d/1hPYA-1X5RBlPzlH-tsdnXjM7wN0wq7FF3BmyigUJOPc/edit?usp=sharing")
``` 

## The "students" dataframe
```{r}
students
```

## Frequency Distribution - Math Scores

First, display the frequency of math scores.

```{r}
table(students$math)
```

## Display Table 2-2 vertically
Note: 31 levels
```{r}
table2.2 <- cbind(Freq = table(students$math))
table2.2
```

## Fix the lack of zeros
We want to show that some of the math scores are zero. (For example, no one scored 59 on the math test.)
```{r}
# Define 42 levels between 19 & 60
table.2.2 <- factor(students$math, levels = c(60:19)) 

# Show the table vertically
table2.2 <- cbind(Freq = table(table.2.2))

# Make this a data frame (It was a vector.)
table2.2 <- as.data.frame(table2.2)

# Add a second column using the row names (60 - 19)
table2.2 <- table2.2 |>
  mutate(table2.2, ScoreX = row.names(table2.2), .before = 1)  # Add ScoreX column
```

## Our new Table 2-2
Frequency Distribution of Scores on the Mathemeatics Test for 52 Students
```{r}
table2.2
``` 

## Grouped Frequency Distribution
As shown on page 31 (one way to approach this)
```{r}
# create a vector called "bins", counting by threes
bins <- seq(17, 62, by=3) 
# NOTE: Starting at 17 lines up with Thorndike

# Then create a vector called "Interval"
Interval <- cut(students$math, bins)
# The "cut()" command divides the range of Table.2.1$Math into intervals and codes the values in x according to which interval they fall.

table(Interval) # Produces a horizontal table
```

## Table 2-3
Grouped Frequency Distribution of Scores from 52 Students on a Math Test Using an Interval of 3
```{r}
# transform() makes a vertical table like Table.2.3
table.2.3 <- transform(table(Interval))
table.2.3
```

## Table 2-3
This time with 60 at the top of the table
```{r}
arrange(table.2.3, desc(Interval)) 
```

## Cumulative Frequency Distribution
page 35
```{r}
# Create data frame of Math scores
# Note: There are much more elegant ways to do this.
table.2.4 <- data.frame(table.2.3) |>
  mutate(Cumulative_Frequency = cumsum(Freq)) |>
  mutate(Cumulative_Percent = round(100*cumsum(Freq)/52))
``` 

## Table 2-4
```{r}
arrange(table.2.4, desc(Interval))
```

## Histograms
```{r}
ggplot(students, aes(x = math, y = after_stat(count))) +
  geom_histogram(binwidth = 3, color = "black", fill = "grey") + 
  theme_classic() +
  labs(x = "Mathematics test scores",
       y = "Frequency") +
  scale_x_continuous(breaks = 17 + c(0:15)*3) +
  scale_y_continuous(breaks = 0 + c(0:6)*2) + 
  ggtitle("Figure 2-1\nHistogram of 52 mathematics scores")
```

## Figure 2-1
```{r, echo=FALSE}
ggplot(students, aes(x = math, y = after_stat(count))) +
  geom_histogram(binwidth = 3, color = "black", fill = "grey") + 
  theme_classic() +
  labs(x = "Mathematics test scores",
       y = "Frequency") +
  scale_x_continuous(breaks = 17 + c(0:15)*3) +
  scale_y_continuous(breaks = 0 + c(0:6)*2) + 
  ggtitle("Figure 2-1\nHistogram of 52 mathematics scores")
```

## Cumulative Frequency Curve
```{r}
# Again, this is one way of approaching it. Not super elegant though
MathByThrees <- data.frame(table.2.4) |>
  mutate(Threes = 20 + c(0:14)*3) 

# Figure 2-3
ggplot(MathByThrees, aes(x=Threes, y=Cumulative_Frequency)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(x = "Math Score",
       y = "Cumulative frequency") +
  ggtitle("Figure 2-3\nCumulative frequency curve") +
  scale_x_continuous(breaks = 18 + c(0:14)*3)
```

## Figure 2-3
```{r, echo=FALSE}
ggplot(MathByThrees, aes(x=Threes, y=Cumulative_Frequency)) +
  geom_line() +
  geom_point() +
  theme_classic() +
  labs(x = "Math Score",
       y = "Cumulative frequency") +
  ggtitle("Figure 2-3\nCumulative frequency curve") +
  scale_x_continuous(breaks = 18 + c(0:14)*3)
```
